---
title: "Advanced workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
has_asreml <- requireNamespace("asreml", quietly = TRUE)
```

```{r setup}
library(InterPlotComp)
```

This vignette covers the three things that turn a working single-trial fit into
a production analysis: a **relationship matrix**, a **covariate** for the
physical cause of interference, and the **multi-environment** model.

It assumes you have read *Basic workflow*.

## Multi-environment trials

A series of environments is prepared through the same function, with
`multi_env = TRUE` and an environment column mapped:

```{r}
met_raw <- sample_met_trial()
table(met_raw$Environment)

met <- prepare_trial_data(
  met_raw,
  map = list(yield = "Yield_t_ha", geno = "Genotype", row = "Row",
             column = "Column", env = "Environment", rep = "Rep",
             block = "Block"),
  multi_env = TRUE
)

attr(met, "field_summary")
```

The four environments have different sizes, different genotype panels and
unequal replication — which is what a real series looks like. Each is padded to
its own rectangle and gets its own residual section:

```{r}
met_nb <- add_neighbours(complete_field_grid(met), axis = "rows")
nlevels(met_nb$data$Env)
```

Design factors are nested inside environment automatically. Block "1" in one
environment is not the same block as "1" in another, and collapsing them is a
common cause of a singular Average Information matrix:

```{r}
head(levels(met_nb$data$RepF))
```

### Choosing the between-environment structure

| `structure` | What it assumes | Parameters |
|---|---|---|
| `"diag"` | No genetic correlation between environments | fewest |
| `"facv"` | Joint factor-analytic covariance over direct *and* competitive effects | most |
| `"separable"` | One 2×2 direct–competition covariance crossed with one environment covariance | middle |
| `"fa"` | Separate FA covariances; direct–competition covariance fixed at zero | middle |

`"facv"` is the model you want if it will fit: it is the only one that lets the
direct–competition covariance differ between environments. `"separable"` is the
usual fallback, costing `3 + E(r + 1)` parameters instead of `2E(r + 1)`.

```{r, eval = has_asreml}
met_fit <- fit_met_model(
  met_nb$data, met_nb$names,
  opts = list(structure = "facv", rank = 1L, spatial = TRUE, nugget = FALSE,
              auto_simplify = TRUE, exact_se = FALSE, cinv_limit = 0L,
              maxit = 25L, workspace = "2gb", compare_baseline = FALSE,
              max_rounds = 10L)
)

met_fit$converged
met_fit$description
```

### Genetic correlations between environments

```{r, eval = has_asreml}
round(met_fit$matrices$direct_cor, 3)
```

These are the correlations that decide whether one selection serves all
environments. Values near 1 mean a genotype that wins in one environment wins
in the others; values near zero mean environment-specific recommendations.

```{r, eval = has_asreml, fig.alt = "Genetic correlation between environments"}
plot_correlation_heatmap(met_fit$matrices$direct_cor,
                         title = "Direct effects")
```

```{r, eval = has_asreml}
met_fit$variance
```

### Stability across environments

```{r, eval = has_asreml, fig.alt = "Pure-stand performance across environments"}
plot_stability(met_fit$values, top_n = 10)
```

Lines that stay parallel are stable genotypes. Lines that cross are crossover
genotype-by-environment interaction, and mean the recommendation depends on the
environment.

## Pedigree and genomic relationships

By default genetic effects are independent between genotypes. Supplying a
relationship matrix replaces that with a known covariance between relatives.
For a competition model this matters twice over: the competitive effect is the
harder of the two to estimate, and it borrows most of its information from
relatives.

Three sources are accepted — a pedigree, a ready-made kinship matrix, or a
marker matrix:

```{r, eval = has_asreml}
ped <- sample_pedigree()
head(ped)

rel <- build_relationship(
  type = "pedigree",
  raw  = ped,
  map  = list(id = "Genotype", sire = "Male_parent", dam = "Female_parent")
)

rel$label
rel$diagnostics[c("n", "n_founders")]
```

A marker matrix goes through VanRaden's method 1, with a small ridge blended in
so the result can be inverted:

```{r, eval = FALSE}
rel <- build_relationship("markers", raw = my_marker_data, blend = 0.01)
```

Pass the result to either fitting function:

```{r, eval = has_asreml}
nb <- add_neighbours(
  complete_field_grid(
    prepare_trial_data(sample_single_trial(),
                       list(yield = "Yield_t_ha", geno = "Genotype",
                            row = "Row", column = "Column", rep = "Rep"))),
  axis = "rows")

rel_fit <- fit_single_model(
  nb$data, nb$names,
  opts = list(structure = "diag", spatial = TRUE, nugget = FALSE,
              auto_simplify = TRUE, exact_se = FALSE, cinv_limit = 0L,
              maxit = 25L, workspace = "1gb", compare_baseline = FALSE,
              max_rounds = 10L, relationship = rel)
)

rel_fit$coverage
```

Individuals in the relationship matrix with no plot in the trial — ancestors,
untested relatives — are kept, and their effects are predicted from their
relatives and flagged:

```{r, eval = has_asreml}
table(rel_fit$genetic$Tested)
```

Every genotype with an *observed* plot must appear in the relationship matrix.
If one does not, the fit stops and names it rather than dropping it silently.

## Adjusting for a covariate

Interference has a physical cause — plant height, canopy width, root vigour.
If you measured it, you can ask how much of the competitive effect it explains.
The covariate enters as a fixed effect twice: once for the plot's own value and
once summed over exactly the neighbours that supply the competitive effects.

```{r, eval = has_asreml}
cov_prepared <- prepare_trial_data(
  sample_single_trial(),
  list(yield = "Yield_t_ha", geno = "Genotype", row = "Row",
       column = "Column", rep = "Rep", covariate = "Plant_height_cm")
)
cov_nb <- add_neighbours(complete_field_grid(cov_prepared), "rows")

cov_fit <- fit_single_model(
  cov_nb$data, cov_nb$names,
  opts = list(structure = "diag", spatial = TRUE, nugget = FALSE,
              auto_simplify = TRUE, exact_se = FALSE, cinv_limit = 0L,
              maxit = 25L, workspace = "1gb", compare_baseline = FALSE,
              max_rounds = 10L, adjust_own_covariate = TRUE)
)

cov_fit$covariate_terms
cov_fit$wald
```

The covariate is centred, so the intercept stays the fitted mean at an average
covariate value and predicted pure-stand yields remain on the scale a breeder
expects. A covariate that is missing for more than half the plots is refused:
adjusting for something mostly absent biases the competitive effects more than
it corrects them.

## The simplification ladder

Both fitting functions take `auto_simplify`. When `TRUE`, a specification that
will not converge is replaced by the next model down a nested ladder, giving up
the least defensible assumption first:

1. a second-order residual process, back to AR1 × AR1;
2. the `us(2)` covariance, reparameterised as `corgh(2)`;
3. the independent nugget variance;
4. the direct–competition covariance, constrained to zero;
5. the block and then replicate variances;
6. the spatial residual — last, because dropping it biases everything else.

```{r, eval = has_asreml}
met_fit$log
```

Always read this log. A result from step 4 is a different scientific claim from
a result at step 1, and the difference belongs in your methods section.

## Reproducible scripting

Nothing above needs the Shiny interface. A complete analysis is:

```{r, eval = FALSE}
library(InterPlotComp)

d  <- prepare_trial_data(read.csv("trial.csv"), map = my_map)
d  <- complete_field_grid(d)
nb <- add_neighbours(d, "rows")
fit <- fit_single_model(nb$data, nb$names, opts = my_opts)

write.csv(fit$genetic, "genotype_values.csv", row.names = FALSE)
```
