---
title: "Basic workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
# Model fitting needs a licensed ASReml-R, which is unavailable on CRAN and on
# most continuous-integration runners. Those chunks are shown but only
# evaluated where the engine is actually present.
has_asreml <- requireNamespace("asreml", quietly = TRUE)
```

```{r setup}
library(InterPlotComp)
```

This vignette runs a complete single-trial analysis, from a raw data frame to a
ranked list of selection candidates. Everything up to the fit itself runs
without ASReml-R.

## 1. The data

The package simulates its worked examples from known parameters, so a fitted
model can be checked against the values that generated the data. The single
trial is 15 field rows by 12 columns, 60 genotypes in an incomplete block
design, with AR1 spatial trend, three failed plots and one field position
physically absent:

```{r}
trial <- sample_single_trial()
head(trial)
dim(trial)
```

Your own data needs one row per plot and, at minimum, four columns: the
response, the genotype, and the two **physical field coordinates**. The
coordinates are not optional and not decorative — they are what makes both the
spatial model and the neighbour lookup possible.

## 2. Preparing the trial

`prepare_trial_data()` validates the file and reshapes it into the internal
analysis layout. You tell it which of your columns is which:

```{r}
prepared <- prepare_trial_data(
  trial,
  map = list(yield  = "Yield_t_ha",
             geno   = "Genotype",
             row    = "Row",
             column = "Column",
             rep    = "Rep",
             block  = "Block")
)
```

It refuses data that cannot support a competition model rather than fitting
something meaningless — too few observed plots, a single genotype, two plots
claiming the same field position, an observed yield with no genotype attached.
It also re-indexes the field coordinates onto a contiguous grid, which matters
more than it sounds: an entirely absent field row would otherwise collapse out
of the factor and make AR1 treat the plots either side of it as adjacent.

A per-environment layout summary comes back attached:

```{r}
attr(prepared, "field_summary")
```

## 3. Completing the field grid

A separable AR1 × AR1 residual needs a complete rectangle. Missing positions
are padded with a missing response; they define the residual covariance layout
and carry no genotype, so they contribute no genetic effect.

```{r}
complete <- complete_field_grid(prepared)
c(observed = nrow(prepared), padded = nrow(complete))
attr(complete, "n_padded")
```

## 4. Attaching neighbours

`add_neighbours()` records which genotype is growing in each adjacent plot.
The axis should match the direction in which plots actually compete — for
single-row plots that is almost always along the field rows.

```{r}
nb <- add_neighbours(complete, axis = "rows")
nb$k        # number of neighbours per plot
nb$label
```

Border plots legitimately have fewer neighbours; their absent neighbour factors
stay `NA` and are absorbed as a zero row in the design matrix.

```{r}
table(nb$data$Neighbour_count)
```

## 5. Fitting the model

```{r, eval = has_asreml}
result <- fit_single_model(
  nb$data, nb$names,
  opts = list(
    structure        = "us",    # unstructured direct-competition covariance
    spatial          = TRUE,    # AR1 x AR1 residual
    nugget           = TRUE,
    auto_simplify    = TRUE,    # fall back if the full model will not converge
    exact_se         = TRUE,
    cinv_limit       = 5000L,
    maxit            = 30L,
    workspace        = "1gb",
    compare_baseline = TRUE,    # test whether competition improves the fit
    max_rounds       = 15L
  )
)

result$converged
result$description
```

`auto_simplify = TRUE` is the safety net. The direct–competition covariance is
the hardest parameter in the model to estimate, and on a small or awkwardly
laid-out trial the full specification may not converge. Rather than failing,
the fit walks down a ladder of nested models, giving up the least defensible
assumption first, and tells you what it gave up:

```{r, eval = has_asreml}
result$log
```

## 6. The variance components

```{r, eval = has_asreml}
result$variance[, c("Component", "Estimate")]
```

The data were simulated with a direct variance of 0.36, a competitive variance
of 0.09 and a direct–competition correlation of −0.55. A negative correlation
is the case that matters to a breeder: the genotypes that yield most in their
own plot are the ones suppressing their neighbours, so part of their apparent
advantage is borrowed and will not survive into a pure stand.

## 7. The genotype table

```{r, eval = has_asreml}
g <- result$genetic
head(g[, c("Genotype", "Direct_effect", "Competition_effect",
           "Pure_stand_effect", "Rank_direct", "Rank_pure_stand",
           "Rank_change")], 10)
```

`Rank_change` is the practical output: how far each genotype moves once its
effect on its neighbours is accounted for.

```{r, eval = has_asreml}
summary(g$Rank_change)
sum(abs(g$Rank_change) >= 5)   # genotypes that move five places or more
```

## 8. Figures

```{r, eval = has_asreml, fig.alt = "Direct effect against competitive effect"}
plot_direct_vs_competition(g, k = result$k)
```

The quadrants separate the decisions. A genotype high on the direct axis and
benign on the competitive axis is an unambiguously good selection. One high on
both is flattered by the trial.

```{r, eval = has_asreml, fig.alt = "Genotypes ranked by predicted pure-stand performance"}
plot_ranking(g, top_n = 20)
```

## 9. Was competition worth modelling?

```{r, eval = has_asreml}
result$comparison$lrt
```

This is a likelihood-ratio test against the identical model with the
competitive effects removed. Both models are iterated to convergence and carry
the same fixed and residual structure, so the test isolates the competition
term. If it is not significant, report the simpler model.

## Doing all of this in the interface

```{r, eval = FALSE}
run_app()
```

The Shiny interface performs exactly this pipeline, with the diagnostics shown
alongside each step and every table and figure exportable.
