---
title: "Interpreting the output"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting the output}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5, dpi = 96)
has_asreml <- requireNamespace("asreml", quietly = TRUE)
```

```{r setup}
library(InterPlotComp)
```

Fitting the model is the easy part. This vignette is about reading it: what
each quantity means, how the pieces relate, and — the section that matters most
— when the answer should not be trusted.

```{r, eval = has_asreml, include = FALSE}
nb <- add_neighbours(
  complete_field_grid(
    prepare_trial_data(sample_single_trial(),
                       list(yield = "Yield_t_ha", geno = "Genotype",
                            row = "Row", column = "Column",
                            rep = "Rep", block = "Block"))),
  axis = "rows")

fit <- fit_single_model(
  nb$data, nb$names,
  opts = list(structure = "us", spatial = TRUE, nugget = TRUE,
              auto_simplify = TRUE, exact_se = TRUE, cinv_limit = 5000L,
              maxit = 30L, workspace = "1gb", compare_baseline = TRUE,
              max_rounds = 15L))
```

## The three genotype effects

For genotype *i* the model estimates a direct effect \(D_i\) and a competitive
effect \(C_i\). With \(k\) neighbours per plot, the pure-stand effect is

$$P_i = D_i + k\,C_i$$

because in a pure stand the genotype is its own neighbour on all \(k\) sides.

| Quantity | Meaning | Sign convention |
|---|---|---|
| `Direct_effect` | What the genotype expresses in its own plot | Higher is better |
| `Competition_effect` | What it does to a neighbouring plot | **Positive is better**: it *raises* its neighbour |
| `Pure_stand_effect` | What it expresses with neighbours like itself | Higher is better |

The sign convention on the competitive effect is the one people get wrong. A
**negative** competitive effect means the genotype *suppresses* its neighbours
— it is the aggressive competitor, and it is flattered by the trial.

```{r, eval = has_asreml}
g <- fit$genetic
head(g[, c("Genotype", "Direct_effect", "Competition_effect",
           "Pure_stand_effect", "Competitor_type")], 8)
```

## The variance components

```{r, eval = has_asreml}
fit$variance[, c("Component", "Estimate", "Interpretation")]
```

The pure-stand variance is not a free parameter; it follows from the other
three:

$$\sigma^2_P = \sigma^2_D + k^2\sigma^2_C + 2k\,\sigma_{DC}$$

```{r, eval = has_asreml}
cmp <- fit$components
c(reported  = cmp$pure,
  recomputed = cmp$direct + fit$k^2 * cmp$competition + 2 * fit$k * cmp$covariance)
```

### The direct–competition correlation

This single number carries most of the biology.

* **Negative** — high-yielding genotypes suppress their neighbours. Part of
  their measured advantage was taken from the plots they are compared against,
  and will not carry into a pure stand or a farmer's field. This is the usual
  finding, and the reason to fit the model at all.
* **Near zero** — the two effects are unrelated; competition adds noise but not
  bias to the ranking.
* **Positive** — unusual; check the field layout and the neighbour axis before
  believing it.

```{r, eval = has_asreml}
cmp$correlation
```

## Reliability and heritability

```{r, eval = has_asreml}
fit$heritability
```

`direct` is generalised heritability on the direct effect; `pure` is the same
quantity on the pure-stand value. **The pure-stand figure is almost always the
lower of the two**, and that is not a defect of the model — it is the honest
statement that the pure-stand value is a derived quantity combining two
estimates and their covariance, so it carries more uncertainty than either.

Per-genotype reliability is on the usual 0–1 scale:

```{r, eval = has_asreml}
summary(g$Reliability_direct)
```

## Does the ranking actually change?

This is the question the whole analysis exists to answer.

```{r, eval = has_asreml}
summary(g$Rank_change)
sum(abs(g$Rank_change) >= 5)
```

```{r, eval = has_asreml, fig.alt = "Rank change once competition is accounted for"}
plot_rank_change(g, top_n = 25)
```

If nothing moves, competition is present but harmless for selection, and you
should say so. If entries move substantially, the trial as recorded was
selecting partly on aggression.

## Was competition worth modelling?

```{r, eval = has_asreml}
fit$comparison$table
fit$comparison$lrt
```

A likelihood-ratio test against the identical model without competitive
effects. Both models are iterated to convergence and share the same fixed and
residual structure, so the test isolates the competition term.

Note that testing a variance component at its boundary makes the ordinary
chi-squared p-value **conservative** — the real p-value is smaller than the one
reported, so a significant result stays significant.

## Model diagnostics

```{r, eval = has_asreml, fig.height = 6, fig.alt = "Residual diagnostics"}
plot_residual_diagnostics(fit$residuals)
```

```{r, eval = has_asreml}
fit$fit_stats
fit$converged
fit$convergence_rounds
```

`convergence_rounds` counts how many restarts were needed after the first fit
ran out of iterations. A model that merely stopped iterating is not safe to
quote, so the fit keeps restarting from the current estimates until ASReml
reports convergence.

## When not to trust the result

Read the fitting log first. Every one of these is reported by the package
itself, in the interface and in the returned object.

**1. The model fell down the simplification ladder.**

```{r, eval = has_asreml}
fit$fallback_used
fit$log
```

If the direct–competition covariance was constrained to zero, you did not
measure it — you assumed it. Report which model you actually fitted.

**2. Too few plots have a complete neighbour set.** The competitive effect is
then estimated mainly from interior plots:

```{r}
diag_nb <- competition_diagnostics(nb$data, nb$names)
diag_nb[c("full_neighbour_pct", "mean_neighbours", "border_plots")]
```

**3. Each genotype sits beside too few distinct neighbours.** If a genotype is
nearly always beside the same neighbour, the direct and competitive effects are
weakly separated and the design cannot tell them apart:

```{r}
diag_nb$mean_distinct_neighbours
```

**4. The genotype panel is small.** Below roughly 15 genotypes the genetic
variances, and especially the direct–competition covariance, are too imprecise
to interpret.

```{r}
competition_warnings(diag_nb)
```

An empty character vector means none of these applies.

**5. A component is at its boundary.** A variance pinned at zero is a
constraint that bound, not an estimate:

```{r, eval = has_asreml}
fit$varcomp[fit$varcomp$At_boundary, c("Component", "Estimate", "Bound")]
```

**6. The competition direction does not match the layout.** `add_neighbours()`
will happily use whichever axis you name. For single-row plots, competition is
along the field rows; using `"columns"` there estimates something real but not
what you meant.

## Reporting

A defensible methods paragraph states:

* the model actually fitted, including anything the ladder gave up
  (`fit$description` and `fit$log`);
* the residual structure (`fit$spec$spatial`, and the AR1 correlations from
  `fit$varcomp`);
* the estimated variance components with standard errors;
* the direct–competition correlation;
* whether competition improved the fit (`fit$comparison$lrt`);
* whether, and by how much, the ranking changed.

```{r, eval = has_asreml}
fit$description
```
