---
title: "Installation and setup"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation and setup}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this package is for

In an unbordered single-row-plot trial, every plot is exposed to whatever is
growing beside it. A vigorous entry takes light, water and nutrients from its
neighbours, so part of the yield recorded against it was taken from the plots
either side. The trial rewards it twice: once for its own performance and once
for depressing the plots it is compared against.

`InterPlotComp` fits the direct–competition model that separates those two
things. For every genotype it estimates

* the **direct effect** — what the genotype expresses in its own plot;
* the **competitive effect** — what it does to a neighbouring plot;
* the **pure-stand value** — what it would express with neighbours like itself,
  which is the quantity selection should act on.

The whole pipeline is available as functions, and as a Shiny interface launched
with `run_app()`.

## Installing the package

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("tolekeno/InterPlotComp")
```

The required packages are pulled in automatically. Several optional packages
each add one feature — interactive figures, higher-quality image export, Excel
output — and every one of them degrades to a working fallback, so a minimal
installation is fully usable:

```{r, eval = FALSE}
remotes::install_github("tolekeno/InterPlotComp", dependencies = TRUE)
```

R 4.1 or later is required.

## Installing ASReml-R

**ASReml-R is commercial software licensed by [VSNi](https://vsni.co.uk). It is
not on CRAN, it is not bundled with this package, and it is not installed by
any of the commands above.** It is the only engine this package has: without a
working, licensed ASReml-R installation no model can be fitted.

To use the modelling functions you must

1. obtain ASReml-R 4.1 or later from VSNi;
2. install it into the same R library that runs `InterPlotComp`;
3. activate its licence, in a plain R session, with
   `asreml::asreml.license.activate()`.

`InterPlotComp` never reads, writes, embeds, stores or transmits a licence key.
It uses whatever ASReml installation and already-activated licence exist in the
R process that runs it.

> **Never commit a licence file or activation key** to a shared or public
> repository, and never bake one into a container image. The package's
> `.gitignore` excludes the usual licence file names as a safety net.

## Checking your installation

Everything that does not fit a model works without ASReml, so you can confirm
the package itself is healthy first:

```{r}
library(InterPlotComp)

trial <- sample_single_trial()
str(trial)
```

To check whether the modelling engine is available:

```{r}
asreml_ready <- requireNamespace("asreml", quietly = TRUE)
asreml_ready
```

That call is also what performs the licence checkout, so `FALSE` means either
"not installed" or "installed but not licensed". The package reports the two
cases separately when you try to fit:

```{r, eval = FALSE}
run_app()   # the interface opens either way and says which applies
```

The interface deliberately opens without ASReml so that an installation or
licensing problem is visible immediately, rather than after you have uploaded a
large file and waited for a fit that was never going to run. Every data panel,
field plan and design summary works; only the fitting buttons are blocked.

## Where to go next

| Vignette | What it covers |
|---|---|
| *Basic workflow* | Preparing a trial and fitting a single-trial model |
| *Advanced workflow* | Relationship matrices, covariates and the MET model |
| *Interpreting the output* | What each number means and when not to trust it |

```{r, eval = FALSE}
vignette("basic-workflow", package = "InterPlotComp")
```
